The Evolution of Phenotypically Invariant Genetic Networks

### 2017-11-20 kalman_walk_paper.tex
discussion edits

The complexity of biological systems has limited our understanding of their function and evolution. Above we outline an approach, a first step, towards untangling this complexity in reference to function and evolution. This methodology borrows successfully applied tools from engineering and aims to synthesize these with the concepts and tools of molecular and evolutionary biology.
Theoretical models in evolution and population genetics often lack the molecular details of physiology or of the genotype-phenotype map. 
Here, we offer a tractable and simple model which includes these missing features. 
Further, we provide, in clear mathematical language, an analytical description of phenomena hitherto discussed verbally and conceptually 
(phenogenetic drift \citep{weiss2000phenogenetic}, developmental system drift \citep{true2001developmental}, biological degeneracy \citep{edelman2001degeneracy}). 
The tractability and relative simplicity of this exposition enables the interested biologist to work out by hand, 
if desired, 
the dynamics of a genetic system, as well as perturbations to the system -- an attribute not likely to be found in less tractable models and simulations.

@@

@@

Our results show that hybrids, under neutral processes, break down as a function of genetic distance, and despite the present model's myriad simplifications and assumptions, 
it may, in part, explain the observed broad patterns of reproductive isolation among diverse phyla \citep{roux2016shedding, hedges2015tree}. \js{maybe also cite the phylogentics lit here? Like the speciation clock stuff.}

Although, as classically formulated, the Dobzhansky-Muller model of hybrid incompatibility is agnostic to the relative importance of neutral versus selective genetic substitutions \citep{coyne1998evolutionary}, \citet{orr2004speciation} remarks that hybrid incompatibility data suggests that ``speciation results from positive Darwinian selection within species'' thus maintaining one of the ``central tenets of the modern synthesis''. 
%In practice it remains unclear whether system drift or other speciation mechanisms predominate, 
%as hybrid incompatibilities due to both selection and conflict have been uncovered and hint at general processes \citep{orr2004speciation}. 
While it is known that divergent selection and genetic conflict, as well as other mechanisms, can precipitate speciation, we have shown here that, surprisingly, neutral drfit may also significantly drive species formation.
This seems reasonable as there is evidence that the set of molecular mechanisms underlying a trait is degenerate, that there is significant regulatory variation segregating within populations, that functionally equivalent systems, even in closesly related species, can be mechanistically divergent \citep{dalal2017transcription}, transcription in hybrids between closely related species with conserved transcriptional patterns can be divergent \citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association}, that gene misregulation can cause dysfunction (by definition), and that numerous cryptic species complexes exist (e.g. sun skinks \citep{barley2013challenge}).

If system drift induced hybrid breakdown occurs as rapidly as our model sugests, even in the absence of postive Darwinian selection, this would imply that species originate not necessarily by means of natural selection \citep{darwin1859origin}. While the incompatibility of any one given network within an organism may be small, we note that organsisms are made up of many different networks, and that the cumulative impact of system drift across all of these may be substantial. Further, even in populations experiencing directional selection, networks underlying conserved parts of the phenotype can still experience neutral drift and harbor incompatibilities.

>>>>

The complexity of biological systems has limited our understanding of their function and evolution. Above we outline an approach, a first step, towards untangling this complexity in reference to function and evolution. This methodology borrows successfully applied tools from engineering and aims to synthesize these with the concepts and tools of molecular and evolutionary biology.
Theoretical models in evolution and population genetics often lack the molecular details of physiology or of the genotype-phenotype map. 
Here, we offer a tractable and simple model which includes these missing features.

Further, we provide, in clear mathematical language, an analytical description of phenomena hitherto discussed verbally and conceptually 
(phenogenetic drift \citep{weiss2000phenogenetic}, developmental system drift \citep{true2001developmental}, biological degeneracy \citep{edelman2001degeneracy}). 
The tractability and relative simplicity of this exposition enables the interested biologist to work out by hand, 
if desired, 
the dynamics of a genetic system, as well as perturbations to the system -- an attribute not likely to be found in less tractable models and simulations.

@@

System drift may be a significant driver of speciation as there is evidence that the set of molecular mechanisms underlying a trait is degenerate, that there is significant regulatory variation segregating within populations, that functionally equivalent systems, even in closesly related species, can be mechanistically divergent \citep{dalal2017transcription}, transcription in hybrids between closely related species with conserved transcriptional patterns can be divergent \citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association}, that gene misregulation can cause dysfunction (by definition), and that numerous cryptic species complexes exist (e.g. sun skinks \citep{barley2013challenge}).

@@

Although, as classically formulated, the Dobzhansky-Muller model of hybrid incompatibility is agnostic to the relative importance of neutral versus selective genetic substitutions \citep{coyne1998evolutionary}, \citet{orr2004speciation} remarks that hybrid incompatibility data suggests that ``speciation results from positive Darwinian selection within species'' thus maintaining one of the ``central tenets of the modern synthesis''. 
%In practice it remains unclear whether system drift or other speciation mechanisms predominate, 
%as hybrid incompatibilities due to both selection and conflict have been uncovered and hint at general processes \citep{orr2004speciation}. 
While it is known that divergent selection and genetic conflict, as well as other mechanisms, can precipitate speciation, we have shown here that, surprisingly, neutral drfit may also significantly drive species formation.

If system drift induced hybrid breakdown occurs as rapidly as our model sugests, even in the absence of postive Darwinian selection, this would imply that species originate not necessarily by means of natural selection \citep{darwin1859origin}. While the incompatibility of any one given network within an organism may be small, we note that organsisms are made up of many different networks, and that the cumulative impact of system drift across all of these may be substantial. Further, even in populations experiencing directional selection, networks underlying conserved parts of the phenotype can still experience neutral drift and harbor incompatibilities. 
Our results show that hybrids, under neutral processes, break down as a function of genetic distance, and despite the present model's myriad simplifications and assumptions, 
it may, in part, explain the observed broad patterns of reproductive isolation among diverse phyla \citep{roux2016shedding, hedges2015tree}. \js{maybe also cite the phylogentics lit here? Like the speciation clock stuff.}
